\section{Introduction}
\label{sec:introduction}
Housing corporations increasingly rely on digital image archives to document the condition of rental properties. Inspectors manually compare groups of lease images to identify issues such as damage, alterations, or wear and tear. This process is slow, subjective, and influenced by variations in lighting, camera angle, and device quality. As the volume of archived material grows, scalable and consistent support becomes relevant to housing organizations and inspectors \cite{Kang_Kim_Kim_2025}.

Deep learning has shown strong performance across visual inspection domains, including infrastructure monitoring, medical imaging, and anomaly detection \cite{Cheng_Huang_Li_Lyu_Xu_Zhao_Zhao_Xiang_2024}. Within supervised object detection, one-stage architectures such as YOLOv8 provide a fast, stable, and widely adopted baseline. YOLOv8 is suitable for this project because it can be trained on small–medium datasets, supports structured categories, and offers reproducible results with limited compute \cite{Gonçalves_Santos_2023}.

In parallel, multimodal foundation models have emerged. Models such as LLaVA-1.6 are pretrained on large-scale image–text corpora and can perform open-vocabulary visual reasoning and classification without task-specific labels. These models demonstrate strong generalization across domains and increasingly appear in inspection-oriented research, including industrial anomaly detection and indoor defect description \cite{Jiang_Li_Deng_Liu_Gao_Zhou_Li_Wang_Zheng_2025, Wen_Chen_2024}. Due to their broad pretraining, foundation models may handle indoor variability—lighting, viewpoint changes, clutter—without additional fine-tuning.

Although prior work shows that both supervised detectors and foundation models can detect defects or anomalies, these model families are almost always evaluated separately, using different datasets, metrics, and tasks. Direct empirical comparison between supervised object detection networks and multimodal foundation models under identical real-world visual inspection conditions remains missing. As a result, their relative robustness, generalization capacity, and suitability for structured inspection scenarios are not yet understood.

Furthermore, the suitability of foundation models for structured lease-housing image assessment—where inspection-relevant categories must be consistently identified and, where possible, localized—has not been systematically evaluated. This represents a specific application gap, especially since indoor housing environments contain high visual variability and task-specific semantics not present in generic pretraining datasets.

This thesis addresses these gaps by comparing a YOLOv8 object-detection model with the multimodal foundation model LLaVA-1.6, evaluating their ability to detect and classify inspection-relevant categories in groups of lease housing images. By evaluating both approaches on identical datasets, metrics, and inspection criteria, this work aims to quantify their performance, robustness, and practical utility for supporting housing corporations in preparing and prioritizing property inspections.

Two core gaps emerge:
\begin{enumerate}
    \item \textbf{Scientific gap:} Direct comparison between supervised object detectors and multimodal foundation models is missing. Existing studies evaluate each model type independently, on unrelated datasets, making their relative robustness and generalization unclear \cite{jiang_mmad_2025, ni_towards_2026, wen_autonomous_2024}.
    \item \textbf{Application gap:} No study has examined multimodal foundation models for structured lease housing inspection, where damage, wear, and alterations must be consistently identified across groups of indoor images \cite{kang_improvement_2025}.
\end{enumerate}

The research question is:

\textbf{Key Research Question}

\emph{To what extent do convolutional object detection networks YOLO based architectures differ in performance from foundational language models (ChatGPT) in detecting and classifying predefined inspection relevant changes within lease housing images under real world variability?}

To answer this question, the following sub-questions are made:

\textbf{Sub-Research Questions}
\begin{itemize}
    \item What is the robustness of convolutional object detection networks and foundation vision language models when applied to groups of lease housing images under real world variability (lighting variation, camera angle differences and device heterogeneity)?
    \item How do convolutional detection networks and foundation vision language models compare quantitatively in detecting and classifying inspection relevant changes within groups of lease housing images (precision, recall, F1-score, and error rates) 
    \item How does a predefined baseline similarity or rule based method perform on the same image groups, and how do both model families improve upon this baseline?
\end{itemize}
