\section{Seminal Papers}

This project builds on existing research in automated visual inspection and change detection. Several works inform the scientific foundation of this proposal  across multiple domains.

Cheng\cite{Cheng_Huang_Li_Lyu_Xu_Zhao_Zhao_Xiang_2024} provides an overview of deep learning based change detection methods, Cheng highlights the role of temporal analysis, identifying structural changes in images. In the built environment, Kang\cite{Kang_Kim_Kim_2025} shows how image-based defect detection can accelerate  automated building inspections, while Choi et al. \cite{Choi_2022} show that visual damage detection can be applied to vandalism assessment in architectural settings.

Related work in uncontrolled environments supports the this approach. Sukel \cite{Sukel_Rudinac_Worring_2020} applies object detection to urban maintenance tasks, Santos Gonçalves et al. \cite{Gonçalves_Santos_2023} address robust indoor object detection under a variety of conditions, and Tan et al. \cite{Tan_Zhai_Zhai_2024} demonstrate detection from non-standard camera perspectives. Studies like medical longitudinal change detection \cite{Kamraoui_Mansencal_Manjon_Coupé_2022} and insurance claim damage assessment \cite{Lee_Lee_Park_2024}, further illustrate that deep learning can support visual assessment under world-like variability.

These studies show that modern computer vision models are effective for detecting damage, defects, and visual changes across domains. None address automated comparison of pre- and post-lease images for indoor housing inspections, leaving a clear gap in combining temporal image comparison with practical inspection support, which this thesis aims to address.

\section{Objectives \& Goals}
\label{sec:objectives_and_goals}

\textbf{Project Goal} \\
This research an Siamese computer vision system that detects and classifies inspection findings like damage or wear-and-tear, from uploaded images. How the system will produce an inspection summary  and attaches evidence per finding. It research how to take the previous inspection report to compare outcomes and highlight findings, while tolerating non identical viewpoints.

\textbf{Dataset Construction and Data Sources} \\
We construct a dataset combining the Building Defects Dataset with asbestos\cite{asbestos}, concrete crack\cite{crack}, and paint-damage datasets\cite{House} from \cite{Roboflow}. Unlike \cite{Kamraoui_Mansencal_Manjon_Coupé_2022}, who employ aligned Siamese medical time-points, our approach compares two unpaired pre- and post- inspection residential image sets, introducing distinct domain and matching challenges. 

\textbf{Objectives}
\begin{itemize}
    \item Identify relevant computer vision methods for visual change detection.
    \item Collect a dataset online inside property images to add complement the existing dataset.
    \item Use a dataset from DGW lease images if it could be shared publicly.
    \item Develop a model to compare and classify visual changes.
    \item Define label categories (like damage, maintenance needs.
    \item Introduce a scoring assessment that indicates urgency or inspection priority.
    \item Evaluate the performance of the model using metrics like confusion matrix.
    \item Reflect on limitations with Siamese modeling.
    \item Introducing a benchmark for identifying damage,  maintenance, wear and tear.
\end{itemize}

\textbf{Extended Goal Breakdown}
\begin{enumerate}
    \item \textbf{Develop an Image-Based Change Detection Model} \\
    \textit{Objective:} Create a computer vision model that scores property images taken before and after leasing periods.
    
    \item \textbf{Classify Types of Visual Changes} \\
    \textit{Objective:} Distinguish between categories such as damage, alterations, wear and tear, and maintenance needs.

    \item \textbf{Introduce a Change Scoring or Severity Mechanism} \\
    \textit{Objective:} Provide a measurable score or classification level to prioritize follow-up actions.

    \item \textbf{Support Inspection Evaluation} \\
    \textit{Objective:} Enable property and housing staff to determine whether a physical inspection is needed before visiting the property.

    \item \textbf{Develop a Proof of Concept} \\
    \textit{Objective:} Build and test a prototype that demonstrates feasibility using sample or real property images.
\end{enumerate}
